
# IE 423: QUALITY ENGINEERING

**Project Part 2**

**11.12.2023**

***Zeynep Sude Aksoy -- 2020402108***

***Hasancan Cebeci -- 2019402036***

**INTRODUCTION**

Project Part 2 explores the application of control charts in pairs
trading, focusing on the hourly time series data of BIST30 stock
indices. In the report highly correlated stock pairs are identified by
using a correlation threshold, realtionship between correlated stocks
are modeled through linear regression, and control charts applied to
monitor residuals for potential trading opportunities.

```{r}
library(tidyverse)
library(lmtest) # For linear regression
library(TTR) # For time-series analysis

# Set the directory where your CSV files are located
data_folder <- "C:/Users/Zeynep Sude Aksoy/Downloads/20180101_20231121_bist30 (1)"
# List all CSV files in the directory
csv_files <- list.files(data_folder, pattern = "*.csv", full.names = TRUE)

# Read and combine all CSV files into a single data frame
combined_data <- lapply(csv_files, read.csv) %>%
  reduce(bind_rows)


# Assuming your combined_data has columns: timestamp, stock_identifier, stock_price
# Pivot the data to have each stock in its own column
reshaped_data <- combined_data %>%
  pivot_wider(names_from = short_name, values_from = price)
head(reshaped_data)
```

In order to identify highly correlated stocks, correlation threshold is
determined as 0.99 and the correlation between each pair of data in
BIST30 is checked.

```{r}
# Calculate correlation matrix for the stocks
correlation_matrix <- cor(reshaped_data[, -1])  # Exclude timestamp column


# Set correlation threshold
correlation_threshold <- 0.99  # Adjust this threshold as needed

# Get the indices where correlation is above the threshold, excluding self-correlations
highly_correlated_pairs <- which(correlation_matrix > correlation_threshold & correlation_matrix < 1, arr.ind = TRUE)

# Create an empty list to store selected pairs
selected_pairs <- list()


# Loop through the correlated pairs to select unique pairs
for (i in 1:nrow(highly_correlated_pairs)) {
  pair <- highly_correlated_pairs[i, ]
  stock1 <- rownames(correlation_matrix)[pair[1]]
  stock2 <- colnames(correlation_matrix)[pair[2]]
  
  # Check if the pair or its reversed version is already selected
  if (!(stock1 %in% unlist(selected_pairs) | stock2 %in% unlist(selected_pairs))) {
    selected_pairs[[length(selected_pairs) + 1]] <- c(stock1, stock2)
  }
}

# Print the selected pairs
selected_pairs
 
```

"GARAN" "AKBNK" is the only pairs that correlation between exceed the
correlation threshold. To select 2 stock pairs, threshold is decreased
to 0.98

```{r}
# Set correlation threshold to different level because we need two pairs
correlation_threshold <- 0.98  # Adjust this threshold as needed

# Get the indices where correlation is above the threshold, excluding self-correlations
highly_correlated_pairs <- which(correlation_matrix > correlation_threshold & correlation_matrix < 1, arr.ind = TRUE)

# Create an empty list to store selected pairs
selected_pairs <- list()

# Loop through the correlated pairs to select unique pairs
for (i in 1:nrow(highly_correlated_pairs)) {
  pair <- highly_correlated_pairs[i, ]
  stock1 <- rownames(correlation_matrix)[pair[1]]
  stock2 <- colnames(correlation_matrix)[pair[2]]
  
  # Check if the pair or its reversed version is already selected
  if (!(stock1 %in% unlist(selected_pairs) | stock2 %in% unlist(selected_pairs))) {
    selected_pairs[[length(selected_pairs) + 1]] <- c(stock1, stock2)
  }
}

# Print the selected pairs
selected_pairs
```

As a result of updated threshold value, 5 different stock pairs exceed
the threshold value and can be used as highly correlated stock pairs for
our analysis.

"KCHOL" "THYAO" and "GARAN" "AKBNK" pairs are selected among 5 pairs.

**Part 1**

In the first part of the project, linear regression modeling is applied
to these highly correlated stock pairs to understand their relationship.
Residuals are calculated under the assumption of constant variance.
Control limits are set, and plots of residuals with control limits are
generated for each pair, aiding in the visualization of deviations from
expected relationships.

Firstly, pair of "KCHOL" or "THYAO" are examined:

```{r}
# Assuming 'combined_data' is your dataset with columns for timestamp, stock_identifier, and stock_price

# Check if values for "KCHOL" or "THYAO" are missing for each timestamp
missing_data_either <- combined_data %>%
  filter(short_name %in% c("KCHOL", "THYAO")) %>%
  group_by(timestamp) %>%
  summarise(missing_count = sum(is.na(price)))

# Filter timestamps where either "KCHOL" or "THYAO" has missing values
timestamps_either_missing <- missing_data_either %>%
  filter(missing_count > 0) %>%
  pull(timestamp)

# Perform linear regression on the selected pair
lm_model_1 <- lm(KCHOL ~ THYAO, data = reshaped_data)

# Get residuals from the linear regression model
residuals_1 <- residuals(lm_model_1)

# Calculate standard deviation of residuals
residuals_sd_1 <- sd(residuals_1)

# Calculate upper and lower control limits (e.g., 3 standard deviations from mean)
upper_limit <- mean(residuals_1) + 3 * residuals_sd_1
lower_limit <- mean(residuals_1) - 3 * residuals_sd_1

# changing format of timestamp from ‘character’ to “POSIXct" "POSIXt" to plot the residuals.
new_format_timestamp = as.POSIXct(reshaped_data$timestamp, format = "%Y-%m-%d %H:%M:%S")


# Create a data frame for plotting
plot_data <- data.frame(timestamp = new_format_timestamp, residuals = residuals_1)
plot_data
# Create a ggplot
ggplot(plot_data, aes(x = timestamp, y = residuals)) +
  geom_line() +
  labs(x = "Timestamp", y = "Residuals", title = "Residuals Plot with Control Limits") +
  geom_hline(yintercept = c(upper_limit, lower_limit), linetype = "dashed", color = c("red", "blue")) +
  scale_x_datetime(date_breaks = "1 month", date_labels = "%b %Y") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  annotate("text", x = as.POSIXct(max(plot_data$timestamp)), y = upper_limit, label = "Upper Limit", hjust = 1.1, color = "red") +
  annotate("text", x = as.POSIXct(max(plot_data$timestamp)), y = lower_limit, label = "Lower Limit", hjust = 1.1, color = "blue") +
  theme(legend.position = "bottom") +
  guides(col = guide_legend(title = "Control Limits"))

```

By monitoring residuals through control chart, it can be seen that there
are residuals that move beyond the control limits. For the HCHOL \~
THYAO pair, according to the control chart, there are residuals exceed
upper control limit between: May 2023-June 2023 and Oct 2023- Dec 2023
and there are residuals exceed lower control limit between: Aug
2023-Sept 2023 and Dec 2022- Jan 2023

In pairs trading with control charts, residuals beyond control limits
signify a potential trading opportunity.

The interpretation of residuals exceeding the control limit as follows:

**Residuals exceeding the upper control limit** suggest that the actual
relationship between the paired stocks deviates significantly from the
expected relationship based on historical data. This could imply that
one stock is outperforming the other beyond what is statistically
expected. In pairs trading, this may trigger a signal to sell the
outperforming stock (short position) and buy the underperforming stock
(long position). The expectation is that the relationship will revert to
its historical mean.

**Residuals falling below the lower control limit** indicate a potential
opportunity where one stock is underperforming more than expected
relative to the other. In response, a pairs trader might initiate a long
position in the underperforming stock and a short position in the
outperforming stock, anticipating a reversion to the mean relationship.

A trading simulation is set up based on the control chart signals and
quantify the potential gains over a specified period, using a
predetermined amount of trading capital which is "1000000". Moreover
shares to trade is pre-determined which is the amount of stock will be
traded according to the relative performance of the paired securities.
For each time period, the action of either "SELL KCHOL/BUY THYAO" or
"BUY KHCOL/SELL THYAO" or "HOLD Both" are determined according to the
residuals position in control chart. At the end of all these trading
decisions, the change in initial capital can be observed.

```{r}
# Define initial trading capital
initial_capital <- 1000000  

# Define the amount of shares to buy or sell based on the trading signal
shares_to_trade <- 1  # You can adjust this value

# Create an empty list to store trade information
trade_list <- list()

# Function to execute trades based on control chart signals
execute_trades <- function(timestamp1, stock1, stock2, residuals, upper_limit, lower_limit) {
  trade_info <- NULL  # Define trade_info within the function
  if (residuals > upper_limit) {
    # Sell stock1 and buy stock2
    action <- "SELL/BUY"
    capital_change <- (stock1 - stock2) * shares_to_trade
  } else if (residuals < lower_limit) {
    # Buy stock1 and sell stock2
    action <- "BUY/SELL"
    capital_change <- -(stock1 - stock2) * shares_to_trade
  } else {
    # No trade
    action <- "HOLD"
    capital_change <- 0
  }
  # Update trade_info
  initial_capital <<- initial_capital + capital_change
  trade_info <- list(timestamp1, stock1, stock2, action, capital_change)
  # Store trade information in the list
  trade_list[[length(trade_list) + 1]] <<- trade_info
}

for (i in 1:length(residuals_1)) {
  execute_trades(reshaped_data$timestamp[i], reshaped_data$KCHOL[i], reshaped_data$THYAO[i], residuals_1[i], upper_limit, lower_limit)
}

# Create the data frame from the list
trades_df1 <- as.data.frame(do.call(rbind, trade_list))
colnames(trades_df1) <- c("Timestamp", "KCHOL", "THYAO", "Action", "CapitalChange")
trades_df1 <- subset(trades_df1, Action != "HOLD")
knitr::kable(trades_df1)
final_capital_1 <- initial_capital
final_capital_1
```

In the output, the points exceeding the control limits coincide with the
intervals that we visually indicate as residuals using the Control chart
graphs. The initial capital of 1000000 becomes 987635.6 as a result of
Pairs Trading Strategy.

Secondly, pair of "GARAN" or "AKBNK" are examined:

```{r}
#For second pair (GARAN and AKBNK)

# Assuming 'combined_data' is your dataset with columns for timestamp, stock_identifier, and stock_price

# Check if values for "GARAN" or "AKBNK" are missing for each timestamp
missing_data_either <- combined_data %>%
  filter(short_name %in% c("GARAN", "AKBNK")) %>%
  group_by(timestamp) %>%
  summarise(missing_count = sum(is.na(price)))



# Perform linear regression on the selected pair
lm_model_2 <- lm(GARAN ~ AKBNK, data = reshaped_data)

# Get residuals from the linear regression model
residuals_2 <- residuals(lm_model_2)


# Calculate standard deviation of residuals
residuals_sd_2 <- sd(residuals_2)

# Calculate upper and lower control limits (e.g., 2 standard deviations from mean)
upper_limit_2 <- mean(residuals_2) + 3 * residuals_sd_2
lower_limit_2 <- mean(residuals_2) - 3 * residuals_sd_2

# Create a data frame for plotting
plot_data_2 <- data.frame(timestamp = new_format_timestamp, residuals = residuals_2)

# Create a ggplot
ggplot(plot_data_2, aes(x = timestamp, y = residuals)) +
  geom_line() +
  labs(x = "Timestamp", y = "Residuals", title = "Residuals Plot with Control Limits") +
  geom_hline(yintercept = c(upper_limit_2, lower_limit_2), linetype = "dashed", color = c("red", "blue")) +
  scale_x_datetime(date_breaks = "1 month", date_labels = "%b %Y") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  annotate("text", x = as.POSIXct(max(plot_data$timestamp)), y = upper_limit, label = "Upper Limit", hjust = 1.1, color = "red") +
  annotate("text", x = as.POSIXct(max(plot_data$timestamp)), y = lower_limit, label = "Lower Limit", hjust = 1.1, color = "blue") +
  theme(legend.position = "bottom") +
  guides(col = guide_legend(title = "Control Limits"))
```

By monitoring residuals through control chart, it can be seen that there
are residuals that move beyond the control limits. For the KCHOL \~
THYAO pair, according to the control chart, there are residuals exceed
upper control limit between: Aug 2023- Sept 2023 and Aug 2022- Sept 2022

The same procedure is followed to implement the pairs trading strategy

```{r}
trade_list <- list()
initial_capital <- 1000000
# Apply trading simulation for the second pair
for (i in 1:length(residuals_2)) {
  execute_trades(reshaped_data$timestamp[i], reshaped_data$GARAN[i], reshaped_data$AKBNK[i], residuals_2[i], upper_limit_2, lower_limit_2)
}

# Create the data frame from the list for the second pair
trades_df2 <- as.data.frame(do.call(rbind, trade_list))
colnames(trades_df2) <- c("Timestamp", "GARAN", "AKBNK", "Action", "CapitalChange")
trades_df2 <- subset(trades_df2, Action != "HOLD")
knitr::kable(trades_df2)
final_capital_2 <- initial_capital
final_capital_2
```

Again in the output, the points exceeding the control limits coincide
with the intervals that we visually indicate as residuals using the
Control chart graphs. The initial capital of 1000000 becomes 1003970 as
a result of Pairs Trading Strategy.

**Part 2**

In order to enhance pairs trading strategy by incorporating more
advanced time series analysis methods for modeling the residuals of
selected stock pairs, autoregressive integrated moving average (ARIMA)
models is implemented to residuals. Since time series models can help
capture patterns, trends, or auto correlations in the residuals that may
not be fully addressed by a basic linear regression model, it is
expected to get a better results in terms of the profitability of pairs
trading strategy. After applying ARIMA model to the difference in stock prices, results
are used to redefine the control limits.

```{r}
#TASK 2

#For the pair KCHOL and THYAO:

library(ggplot2)
library(forecast)  # for ARIMA modeling and forecasts


# Function to fit ARIMA model to residuals and calculate new control limits
fit_arima_model <- function(residuals) {
  # Fit ARIMA model
  arima_model <- auto.arima(residuals)  # Adjust order as needed
  # Extract residuals from the ARIMA model
  arima_residuals <- residuals(arima_model)
  
  # Calculate upper and lower control limits (e.g., 3 standard deviations from mean)
  upper_limit <- mean(arima_residuals) + 3 * sd(arima_residuals)
  lower_limit <- mean(arima_residuals) - 3 * sd(arima_residuals)
  
  return(list(arima_model = arima_model, upper_limit = upper_limit, lower_limit = lower_limit))
}

```

In this part, we changed our approach. Since we picked the most correlated, the prices are expected to follow the same trend in their prices. More specifically, their standart deviation, therefore deviation amount from their mean should be close to each other. Hence, their stock price difference should be consistent in a particular time period. Using autocorrelation function to find a trend between their consequtive stock price differences, we are expected to find more explainatory results compared to simple linear regression.

```{r}


# Finding their stock price difference
task2_residuals_1 = reshaped_data$KCHOL - reshaped_data$THYAO


# Apply ARIMA model and redefine control limits for the first pair
arima_results_1 <- fit_arima_model(task2_residuals_1)

# Extract ARIMA model and new control limits
arima_model_1 <- arima_results_1$arima_model
upper_limit_arima_1 <- arima_results_1$upper_limit
lower_limit_arima_1 <- arima_results_1$lower_limit

plot_data_arima1 <- data.frame(
  timestamp = new_format_timestamp,
  residuals = task2_residuals_1,
  upper_limit = upper_limit,
  lower_limit = lower_limit,
  upper_limit_arima = upper_limit_arima_1,
  lower_limit_arima = lower_limit_arima_1
)
# Red lines represent new control limits
ggplot(plot_data_arima1, aes(x = timestamp, y = residuals)) +
  geom_line() +
  geom_ribbon(aes(ymin = lower_limit, ymax = upper_limit), fill = "blue", alpha = 0.3, color = NA) +
  geom_ribbon(aes(ymin = lower_limit_arima_1, ymax = upper_limit_arima_1), fill = "red", alpha = 0.3, color = NA) +
  labs(x = "Timestamp", y = "Residuals", title = "Comparative Control Limits") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  scale_fill_manual(values = c("blue", "red"), name = "Limits", labels = c("Basic (Blue)", "ARIMA (Red)")) +
  guides(fill = guide_legend(title = NULL))
```

As it can be seen from the control chart of "KCHOL" and "THYAO", control
limits becomes narrower compared to those derived from previous model.
This narrowing of control limits can be a result of the ARIMA model
capturing and incorporating the temporal dependencies or patterns
present in our data. Moreover, in Task 1, variance is assumed constant.
The incorporation of autoregressive (AR) and moving average (MA)
components in the ARIMA model can lead to a reduction in the overall
variability of the residuals. This reduction may result in narrower
control limits, reflecting a more precise estimation of the expected
range of values.

Pair trading strategy is applied to revised control limits.

```{r}
trade_list <- list()
initial_capital <- 1000000
# Apply trading simulation for the second pair
for (i in 1:length(task2_residuals_1)) {
  
  execute_trades(reshaped_data$timestamp[i], reshaped_data$KCHOL[i], reshaped_data$THYAO[i], task2_residuals_1[i], upper_limit_arima_1, lower_limit_arima_1)
}

# Create the data frame from the list for the second pair
trades_dfar1 <- as.data.frame(do.call(rbind, trade_list))
colnames(trades_dfar1) <- c("Timestamp", "KCHOL", "THYAO", "Action", "CapitalChange")
trades_dfar1 <- subset(trades_dfar1, Action != "HOLD")
knitr::kable(trades_dfar1)
final_capital_arima1 <- initial_capital
final_capital_arima1
```

As expected, more residuals beyond control limits are detected hence
more trading decisions are made. The initial capital of 1000000 becomes
1263078 as a result of Pairs Trading Strategy which is higher than the
result ( 987635.6 )in Task 1.

Secondly, pair of "GARAN" or "AKBNK" are examined:

```{r}

# Finding their stock price difference
task2_residuals_2 = reshaped_data$GARAN - reshaped_data$AKBNK


# Apply ARIMA model and redefine control limits for the second pair
arima_results_2 <- fit_arima_model(task2_residuals_2)

# Extract ARIMA model and new control limits
arima_model_2 <- arima_results_2$arima_model
upper_limit_arima_2 <- arima_results_2$upper_limit
lower_limit_arima_2 <- arima_results_2$lower_limit


#### Create a data frame for plotting using 95% prediction interval
plot_data_arima2 <- data.frame(
  timestamp = new_format_timestamp,
  residuals = task2_residuals_2,
  upper_limit = upper_limit_2,
  lower_limit = lower_limit_2,
  upper_limit_arima = upper_limit_arima_2,
  lower_limit_arima = lower_limit_arima_2
)
ggplot(plot_data_arima2, aes(x = timestamp, y = residuals)) +
  geom_line() +
  geom_ribbon(aes(ymin = lower_limit, ymax = upper_limit), fill = "blue", alpha = 0.3, color = NA) +
  geom_ribbon(aes(ymin = lower_limit_arima_2, ymax = upper_limit_arima_2), fill = "red", alpha = 0.3, color = NA) +
  labs(x = "Timestamp", y = "Residuals", title = "Comparative Control Limits") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  scale_fill_manual(values = c("blue", "red"), name = "Limits", labels = c("Basic (Blue)", "ARIMA (Red)")) +
  guides(fill = guide_legend(title = NULL))
```

The same procedure is followed to implement the pairs trading strategy

```{r}
trade_list <- list()
initial_capital <- 1000000
# Apply trading simulation for the second pair
for (i in 1:length(task2_residuals_2)) {
  execute_trades(reshaped_data$timestamp[i], reshaped_data$GARAN[i], reshaped_data$AKBNK[i], task2_residuals_2[i], upper_limit_arima_2, lower_limit_arima_2)
}

# Create the data frame from the list for the second pair
trades_dfar2 <- as.data.frame(do.call(rbind, trade_list))
colnames(trades_dfar2) <- c("Timestamp", "KCHOL", "THYAO", "Action", "CapitalChange")
trades_dfar2 <- subset(trades_dfar2, Action != "HOLD")
knitr::kable(trades_dfar2)
final_capital_arima2 <- initial_capital
final_capital_arima2
```

As in the previous stock pairs, more residuals beyond control limits are
detected hence more trading decisions are made. The initial capital of
1000000 becomes 1070224 as a result of Pairs Trading Strategy which is
higher than the result ( 1003970 )in Task 1.


**Conclusion**

Although, Pairs Trading Strategy seems logical and profitable, basic strategy such as simple linear regression on data is not enough to gain significant benefit or even gain any benefit from this strategy. Creating confidence intervals using high standart deviations and means which are obtained using linear regression gives us a small space to trade and earn profit, because most of the time the process is observed between control limits. Therefore, based on findings, our first impression on Pairs Trading is worse than expected. Using linear regression to trade end up in losing 2% of initial assets in first pair, and gainin only 0.39% of inital assets in the second pair. These values quite small compared to the profits made using more complex strategy.

Secondly, to improve our results and understand the data better, we used auto regression and moving average to explain the relative price changes in stock pairs gives us smaller standart deviations and means, indicating our process is more out of control than we first guessed. Small standart deviation in this case shows us the second model fits the data which is used in second case better than the model used in the first case. Accordingly, using a better fitting model, ARIMA in our case, instead of linear regression led to an increase in the initial assets by 26% whereas using linear regression resulted in losing 2% of the initial assets in the first case. For the second case, using ARIMA increased our earnings by 1700%, from 0.39% to 7%. 

Unfortunately, 7% is not a significant profit at a time scale of 6 years. Probably buying government bonds is more profitable than this strategy in most cases. 26% on the other hand, might be an acceptable profit depending on the interest rates. This analysis shows us, more comprehensive strategy of the data might let us gain more profit from this trading strategy. Also, there is a downside of this strategy. This strategy will fail, when a company crush is occured in selected stocks. A lot of money, will pour into crushed company due to its low price. 


